```{=html}
<style>
body {
  background-color: #fff5f8;
}

h1, h2, h3 {
  color: #c2185b;
}

.title {
  color: #ad1457;
}

a {
  color: #d81b60;
}
</style>
```

---
title: "Assignment 2"
author: "Emma Rodriguez and Lauren Chetley"
format: html
editor: visual
---

## Quarto

## Loading in the data

```{r}
library(dataRetrieval)
library(tidyverse)

siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)
```

## Organizing Data

```{r}
# rename columns for clarity
library(date)
library(lubridate)
library(tidyverse)
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()

```

## Monthly Summary Data - January

```{r}
#filter for January Temperatures 
library(janitor)
temp_data <- temp_data |>
  clean_names()

jan_data<-temp_data|>
  filter(month=="Jan")

jan_summary <- jan_data |>
  group_by(site_no, year) |>
  summarize(mean_temp = mean(temp_C, na.rm = TRUE), .groups = "drop")

ggplot(jan_data, aes(x = year, y = temp_c, color = site_no)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c(
    "#4CC9F0", 
    "#F72585", 
    "#66FF00"  
  )) +
  labs(title = "January Stream Temperature Trends at Willamette River Sites",
       x = "Year",
       y = "Temperature (°C)",
       color = "Site Number") +
  theme_minimal()
  

#make some summary statistics 
#make a glm - generalized linear model for the data pulled abor 

```

## Monthly Summary Data - August

This section examines August stream temperatures at three Willamette River sites from 2010 to 2023. August represents late summer conditions, when stream temperatures are typically at their highest and most relevant for aquatic ecosystems. The goal of this analysis is to evaluate whether August temperatures are changing over time and to compare trends across sites.

```{r}

aug_data <- subset(temp_data, month == "Aug")

colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")

temp_data$month <- month(temp_data$Date, label = TRUE)
temp_data$year <- year(temp_data$Date)

aug_data <- subset(temp_data, month == "Aug")

aug_model <- glm(temp_C ~ year * site_no, data = aug_data)

summary(aug_model)


```

```{r}
ggplot(aug_data, aes(x = year, y = temp_C, color = site_no)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("#08306B", "#2171B5", "#6BAED6")) +
  labs(
    title = "August Stream Temperature Trends",
    x = "Year",
    y = "Temperature (°C)",
    color = "Site"
  ) +
  theme_minimal()
```


### Results Summary

Both the model results and the plot show a clear increase in August stream temperatures over time. The positive and highly significant year coefficient (0.13, p \< 0.001) indicates that temperatures are rising by about 0.13°C per year at the baseline site.

The plot supports this pattern, with upward-sloping trend lines across all sites. However, the rate of warming differs slightly between locations. Site 14211010 shows a significantly faster increase in temperature compared to the baseline, while site 14211720 does not differ significantly.

Overall, August temperatures are increasing across the Willamette River sites, with some variation in how quickly different locations are warming.
