```{=html}
<style>
body {
  background-color: #fff5f8;
}

h1, h2, h3 {
  color: #c2185b;
}

.title {
  color: #ad1457;
}

a {
  color: #d81b60;
}
</style>
```

---
title: "Assignment 2"
author: "Lauren Chetley and Emma Rodriguez"
format: html
editor: visual
---

## Data Loading

```{r}
library(dataRetrieval)
library(tidyverse)

siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)
```

## Organizing Data

```{r}
# rename columns for clarity
library(date)
library(lubridate)
library(tidyverse)
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()

```

## Monthly Summary Data - January

This section examines January stream temperatures from 2010–2023 across three Willamette River sites to assess winter temperature patterns and potential trends over time. The aim of this analysis is to identify whether January temperatures are changing across the different river sites.

```{r}


library(janitor)
temp_data <- temp_data |>
  clean_names()

jan_data<-temp_data|>
  filter(month=="Jan")

jan_summary <- jan_data |>
  group_by(site_no, year) |>
  summarize(mean_temp = mean(temp_c, na.rm = TRUE), .groups = "drop")

ggplot(jan_data, aes(x = year, y = temp_c, color = site_no)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c(
    "#4CC9F0", 
    "#F72585", 
    "#66FF00"  
  )) +
  labs(title = "January Stream Temperature Trends at Willamette River Sites",
       x = "Year",
       y = "Temperature (°C)",
       color = "Site Number") +
  theme_minimal()

jan_data <- subset(temp_data, month == "Jan")

colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_c")

temp_data$month <- month(temp_data$Date, label = TRUE)
temp_data$year <- year(temp_data$Date)

jan_data <- subset(temp_data, month == "Jan")
Jan_model <- glm(temp_c ~ year * site_no, data = jan_data)

summary(Jan_model)
  

```

### Results Summary January 

January stream temperatures show a slight upward trend over time, though this is only marginally significant. Temperatures differ across sites, with some sites consistently cooler than others. One site shows a stronger and statistically significant warming trend, suggesting that temperature changes vary by location.

## Monthly Summary Data - August

This section examines August stream temperatures at three Willamette River sites from 2010 to 2023. August represents late summer conditions, when stream temperatures are typically at their highest and most relevant for aquatic ecosystems. The goal of this analysis is to evaluate whether August temperatures are changing over time and to compare trends across sites.

```{r}

aug_data <- subset(temp_data, month == "Aug")

colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")

temp_data$month <- month(temp_data$Date, label = TRUE)
temp_data$year <- year(temp_data$Date)

aug_data <- subset(temp_data, month == "Aug")

aug_model <- glm(temp_C ~ year * site_no, data = aug_data)

summary(aug_model)


```

```{r}

ggplot(aug_data, aes(x = year, y = temp_C, color = site_no)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("#08306B", "#2171B5", "#6BAED6")) +
  labs(
    title = "August Stream Temperature Trends",
    x = "Year",
    y = "Temperature (°C)",
    color = "Site"
  ) +
  theme_minimal()

```

<<<<<<< HEAD
### Results Summary
=======
### Results Summary August
>>>>>>> 3611aa925ec1cb4235d072ce59becfe633d32732

Both the model results and the plot show a clear increase in August stream temperatures over time. The positive and highly significant year coefficient (0.13, p \< 0.001) indicates that temperatures are rising by about 0.13°C per year at the baseline site.

The plot supports this pattern, with upward-sloping trend lines across all sites. However, the rate of warming differs slightly between locations. Site 14211010 shows a significantly faster increase in temperature compared to the baseline, while site 14211720 does not differ significantly.

Overall, August temperatures are increasing across the Willamette River sites, with some variation in how quickly different locations are warming.
<<<<<<< HEAD
=======

## Conclusion

Overall, stream temperatures in both August and January show increasing trends over time, though the warming is much stronger and more consistent in August. While all sites generally follow similar patterns, the rate of warming varies by location, with some sites experiencing faster temperature increases than others. These results suggest that stream temperatures are rising across the Willamette River system, which could affect water conditions and aquatic life.
>>>>>>> 3611aa925ec1cb4235d072ce59becfe633d32732
