```{=html}
<style>
body {
  background-color: #fff5f8;
}

h1, h2, h3 {
  color: #c2185b;
}

.title {
  color: #ad1457;
}

a {
  color: #d81b60;
}
</style>
```

---
title: "Assignment 2"
author: "Emma Rodriguez and Lauren Chetley"
format: html
editor: visual
---

## Quarto

## Loading in the data

```{r}
library(dataRetrieval)
library(tidyverse)

siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)
```

## Organizing Data

```{r}
# rename columns for clarity
library(date)
library(lubridate)
library(tidyverse)
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()

```

## Monthly Summary Data - January

This section examines January stream temperatures from 2010–2023 across three Willamette River sites to assess winter temperature patterns and potential trends over time. The aim of this analysis is to identify whether January temperatures are changing across the different river sites.

```{r}


library(janitor)
temp_data <- temp_data |>
  clean_names()

jan_data<-temp_data|>
  filter(month=="Jan")

jan_summary <- jan_data |>
  group_by(site_no, year) |>
  summarize(mean_temp = mean(temp_C, na.rm = TRUE), .groups = "drop")

ggplot(jan_data, aes(x = year, y = temp_c, color = site_no)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c(
    "#4CC9F0", 
    "#F72585", 
    "#66FF00"  
  )) +
  labs(title = "January Stream Temperature Trends at Willamette River Sites",
       x = "Year",
       y = "Temperature (°C)",
       color = "Site Number") +
  theme_minimal()

jan_data <- subset(temp_data, month == "Jan")

colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_c")

temp_data$month <- month(temp_data$Date, label = TRUE)
temp_data$year <- year(temp_data$Date)

jan_data <- subset(temp_data, month == "Jan")
Jan_model <- glm(temp_c ~ year * site_no, data = jan_data)

summary(Jan_model)
  

```

January stream temperatures show a slight upward trend over time, though this is only marginally significant. Temperatures differ across sites, with some sites consistently cooler than others. One site shows a stronger and statistically significant warming trend, suggesting that temperature changes vary by location.
